Computing properties of independent sets by generic programming tensor networks \thanks{\funding{...}}

### 2022-01-08 paper/paper.tex
revision of section 7

\subsubsection{Number of independent sets and hard-square entropy constant}
We compute the counting of all independent sets for graphs shown in Figure~\ref{fig:lattices}, where vertices are all placed on square lattices with lattice dimensions $L \times L$.
The graphs include: the square grid graphs, denoted as SG; the square grid graphs with a filling factor $p$, denoted as SG$(p)$, where $\lfloor pL^{2} \rceil$ square grids are occupied with vertices;
the King's graphs, denoted as K; the King's graphs with a filling factor $p$, denoted as K$(p)$. 
The number of independent sets for square grid graphs of size $L \times L$ form a well-known integer sequence (\href{https://oeis.org/A006506}{OEIS A006506}), which is thought as a two-dimensional generalization of the Fibonacci numbers.
We computed the integer sequences for $L=38$ and $L=39$, which is not known before to our knowledge.
In the computation, we have used the finite field algebra trick for contracting arbitrarily high precision integer tensor networks.
We also computed the quantity $F(L,L)^{1/\lfloor pL^2\rceil}$ as a function of grid size $L$ for these lattices,
where $F(L,L)$ is the number of independent sets of a given grid size.
The results are shown in \Fig{fig:hardsquare}.
Our result for $p=1$ matches those in the literature, and the lattices with hole has much larger average value,
which has a phsical meaning of the hard lattices with holes have much larger mean entropy than those without holes at the high temperature limit.

\begin{figure}[t] 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{lattices.pdf}
    \caption{The types of graphs used in the benchmark and case studies.
    The lattice dimensions are $L\times L$. (a) Square grid graph denoted as SG. (b) Square grid graph with a filling factor $p=0.8$, denoted as SG(0.8).
    (c) King's graph denoted as K. (d) King's graph with a filling factor $p=0.8$, denoted as K(0.8).}
    \label{fig:lattices}
\end{figure}

\begin{figure}[t] 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig5.pdf}
    \caption{Mean entropy for lattices defined in \Fig{fig:lattices}.
    We sampled $1000$ instances for $p=0.8$ lattices, and the error bar is too small to be visible.
    The horizontal black dashed lines are for $\lim_{L\rightarrow \infty} F(L,L)^{1/\lfloor pL^2 \rceil}$.
    }
    \label{fig:hardsquare}
\end{figure}

\subsubsection{Computing the overlap gap property}

With the ability to enumerate configurations, we can analyse the landscape of the target problem, like the overlap gap property~\cite{Gamarnik2013, Gamarnik2019}.
We enumeate all ISs of size $\geq\alpha(G)-1$ for two random 3 regular graph instances of size 100, and show the Hamming distance statistics in \Fig{fig:hamming}.
The multiple peak structure indicates disconnected clusters in the configuration space.
\begin{figure} 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig3.pdf}
    \caption{The distribution of Hamming distances between two independent sets of size $\geq\alpha(G)-1$ for two random three regular graph instances of size $100$.}
    \label{fig:hamming}
\end{figure}

\section{Discussion and conclusion}
In this paper, we introduce an abstract algebra formalism to compute properties of independent sets.
The properties include the MIS size, number of independent set of a given size, and enumeration of independent sets of a give size.
For each property, we design a algebra being a commutative semiring, and map the property computation to the contraction of a tensor network with such element type.
The data types introduced in the main text can be diagramatically represented as the following, where we use overlap to represent two algebras can be combined to create a new algebra.
\centerline{\begin{tikzpicture}[]
    \scriptsize
    \node[ellipse,draw,minimum width=2cm, minimum height=1cm,inner sep=0pt] at (0, 2.5) (R) {$\mathbbm{R}$};
    % set and set sampler
    \node[ellipse,draw,minimum width=5cm, minimum height=3cm,inner sep=0pt] at (0, 0) (SN) {};
    \node[below of=1] at (SN) {Set};
    \node[ellipse,draw,minimum width=4cm, minimum height=1.5cm,inner sep=0pt] at (-0.5, 0) (S1) {\hspace{1.5cm}Bit string};
    \def\dx{-3};
    \node[ellipse,draw,minimum width=5cm, minimum height=3cm,inner sep=0pt] at (\dx, 0) (PN) {\hspace{-3.5cm}Polynomial};
    \node[ellipse,draw,minimum width=3.5cm, minimum height=1.2cm,inner sep=0pt] at (\dx, 0.7) (P1) {\hspace{-0.7cm}Highest order};
    \node[ellipse,draw,minimum width=3.5cm, minimum height=0.7cm,inner sep=0pt] at (\dx, -0.4) (P2) {\hspace{-0.7cm}2nd highest order};
    \node[ellipse,draw,minimum width=0.8cm, minimum height=0.3cm,inner sep=0pt] at (\dx, 1.0) (T) {};
    \node at (\dx, -1.2) {$\ldots$};
    \node[above of=2] at (T) (textT) {Tropical};
    \draw[black,-latex] (textT) -- (T);
\end{tikzpicture}}
%We derived the backward rule for tropical tensor network to bound the search of solution space.
%Although many of these properties are global, we can encode them to different tensor element types as commutative semirings.
Except for real numbers, we have polynomials and truncated polynomials algebras. We combine them with set algebra and bit string algebra for set enumeration and sampling.
We call a tensor network with generic tensor element type a generic tensor network, and its power is not limited to solving the independent set properties.
In \App{app:otherproblems}, we show how to map the maximal independent set problem, matching problem, $k$-colouring problem, max cut problem and set packing problem to tensor networks.
Moreover, since the independence polynomial is closely related to the matching polynomial~\cite{Levit2005}, the clique polynomial~\cite{Hoede1994}, and the vertex cover polynomial~\cite{Akbari2013},
our algorithm to compute the independence polynomial can also be used to compute these graph polynomials.
%Some properties of the independence polynomial, such as the uni-modality, log-concavity, and roots of the polynomial, are well studied in the literature for special classes of graphs~\cite{Levit2005}.
%We hope our algorithms to calculate the independence polynomial can also help further research in these directions.
We show some of the Julia language implementations in Appendix~\ref{sec:technical} and you will find it surprisingly short.
A complete implementation can be found in our Github repository~\cite{GraphTensorNetworks}.
%What we need to do is just defining two operations $\oplus$ and $\odot$ and two special elements $\mymathbb{0}$ and $\mymathbb{1}$.
%The style that we program is called generic programming,

>>>>

\subsubsection{Number of independent sets and entropy constant for some hardcore lattice gases}\label{sec:entropy}
We compute the counting of all independent sets for graphs shown in Figure~\ref{fig:lattices}, where vertices are all placed on square lattices with lattice dimensions $L \times L$.
The types of graphs include: the square grid graphs (Figure~\ref{fig:lattices}(a)); the square grid graphs with a filling factor $p=0.8$, which means $\lfloor pL^{2} \rceil$ square grids are occupied with vertices  (Figure~\ref{fig:lattices}(b));
the King's graphs  (Figure~\ref{fig:lattices}(c)); the King's graphs with a filling factor $p = 0.8$  (Figure~\ref{fig:lattices}(d)).
The number of independent sets for square grid graphs of size $L \times L$ form a well-known integer sequence (\href{https://oeis.org/A006506}{OEIS A006506}), which is thought as a two-dimensional generalization of the Fibonacci numbers.
We computed the integer sequence for $L=38$ and $L=39$, which, to the best of our knowledge, is not known before.
In the computation, we used the finite-field algebra for contracting arbitrarily high precision integer tensor networks.

A useful number that can be computed using the number of independent sets is the entropy constant for the hardcore lattice gases on these graphs. For the square grid graphs, it is called the \textit{hard square entropy constant} (\href{https://oeis.org/A085850}{OEIS A085850}), which is defined as $\lim_{L\rightarrow \infty} F(L,L)^{1/L^2}$, where $F(L,L)$ is the number of independent sets of a given lattice dimensions $L \times L$. This quantity arises in statistical mechanics of hard-square lattice gases~\cite{Baxter1980, Pearce1988} and is used to understand phase transitions for these systems. This entropy constant is not known to have an exact representation, but it is accurately known in many digits. Similarly, we can define entropy constants for other lattice gases. In \Fig{fig:hardsquare}, we look at how $F(L,L)^{1/\lfloor pL^2\rceil}$ scales as a function of the grid size $L$ for all types of graphs shown in Figure~\ref{fig:lattices}. Our results match the known results for the non-disordered square grid and King's graphs. For disordered square grid and King's graphs with a filling factor $p=0.8$, we randomly sample 1000 graph instances. To our knowledge, the entropy constants for these disordered graphs have not been studied before. Interestingly, the variations due to different random instances are negligible for this entropy quantity. 
% The lattices with holes have a much larger average value, which has an intuitive physical meaning that the hardcore lattices with holes have a much larger mean entropy than those without holes at the high temperature limit.

\begin{figure}[t] 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{lattices.pdf}
    \caption{The types of graphs used in the case study in section \ref{sec:entropy}.
    The lattice dimensions are $L\times L$. (a) Square grid graphs. (b) Square grid graphs with a filling factor $p=0.8$.
    (c) King's graphs. (d) King's graphs with a filling factor $p=0.8$.}
    \label{fig:lattices}
\end{figure}

\begin{figure}[t] 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig5.pdf}
    \caption{Mean entropy for lattice gases on graphs defined in \Fig{fig:lattices}.
    We sampled $1000$ instances for $p=0.8$ lattices and the error bar is too small to be visible.
    The horizontal black dashed lines are for $\lim_{L\rightarrow \infty} F(L,L)^{1/L^2}$ for the corresponding non-disordered square grid and King's graphs.
    }
    \label{fig:hardsquare}
\end{figure}

\subsubsection{The overlap gap property}
With the tool to enumerate configurations, one can try to understand the structure of the independent set configuration space, such as the optimization landscape for finding the MISs and the geometry of the solution space. One of the known barriers for finding the MIS is the so-called overlap gap property~\cite{Gamarnik2013, Gamarnik2019}. Intuitively, if the overlap gap property is present, it means every two large independent sets either have a significant intersection or very small intersection; it implies that large independent sets are clustered together. This clustering property has been used to prove the limitations of local algorithms in finding the MISs~\cite{Gamarnik2013, Gamarnik2019}. Here, we use the computed configurations of large independent sets to investigate the presence or absence of the overlap gap property. In particular, in \Fig{fig:hamming}, we enumerated all ISs of size $\geq\alpha(G)-1$ for two random three regular graph instances of size 100 and show the pairwise Hamming distance statistics for the enumerated configurations. In \Fig{fig:hamming}(a), we observed the multiple peak structure for the pairwise Hamming distance for one of the graphs. This indicates the presence of the overlap gap property and disconnected clusters exist in the configuration space for large independent sets. We expect our numerical tool can be used to understand this phenomenon better and to further investigate the graph properties and the geometry of the configuration spaces for a variety of graph instances.
\begin{figure} 
    \centering
    \includegraphics[width=\textwidth, trim={0cm 0cm 0cm 0cm}, clip]{figures/fig3.pdf}
    \caption{The distribution of pairwise Hamming distances between every two independent sets of size $\geq\alpha(G)-1$ for two random three regular graph instances of size $100$.}
    \label{fig:hamming}
\end{figure}

\section{Discussion and conclusion}
In this paper, we introduce an abstract algebra formalism to compute properties of independent sets.
The properties include the MIS size, number of independent set of a given size, and enumeration of independent sets of a give size.
For each property, we design a algebra being a commutative semiring, and map the property computation to the contraction of a tensor network with such element type.
The data types introduced in the main text can be diagramatically represented as the following, where we use overlap to represent two algebras can be combined to create a new algebra.

\centerline{\begin{tikzpicture}[]
    \scriptsize
    \node[ellipse,draw,minimum width=2cm, minimum height=1cm,inner sep=0pt] at (0, 2.5) (R) {$\mathbbm{R}$};
    % set and set sampler
    \node[ellipse,draw,minimum width=5cm, minimum height=3cm,inner sep=0pt] at (0, 0) (SN) {};
    \node[below of=1] at (SN) {Set};
    \node[ellipse,draw,minimum width=4cm, minimum height=1.5cm,inner sep=0pt] at (-0.5, 0) (S1) {\hspace{1.5cm}Bit string};
    \def\dx{-3};
    \node[ellipse,draw,minimum width=5cm, minimum height=3cm,inner sep=0pt] at (\dx, 0) (PN) {\hspace{-3.5cm}Polynomial};
    \node[ellipse,draw,minimum width=3.5cm, minimum height=1.2cm,inner sep=0pt] at (\dx, 0.7) (P1) {\hspace{-0.7cm}Highest order};
    \node[ellipse,draw,minimum width=3.5cm, minimum height=0.7cm,inner sep=0pt] at (\dx, -0.4) (P2) {\hspace{-0.7cm}2nd highest order};
    \node[ellipse,draw,minimum width=0.8cm, minimum height=0.3cm,inner sep=0pt] at (\dx, 1.0) (T) {};
    \node at (\dx, -1.2) {$\ldots$};
    \node[above of=2] at (T) (textT) {Tropical};
    \draw[black,-latex] (textT) -- (T);
\end{tikzpicture}}

%We derived the backward rule for tropical tensor network to bound the search of solution space.
%Although many of these properties are global, we can encode them to different tensor element types as commutative semirings.
Except for real numbers, we have polynomials and truncated polynomials algebras. We combine them with set algebra and bit string algebra for set enumeration and sampling.
We call a tensor network with generic tensor element type a generic tensor network, and its power is not limited to solving the independent set properties.
In \App{app:otherproblems}, we show how to map the maximal independent set problem, matching problem, $k$-colouring problem, max cut problem and set packing problem to tensor networks.
Moreover, since the independence polynomial is closely related to the matching polynomial~\cite{Levit2005}, the clique polynomial~\cite{Hoede1994}, and the vertex cover polynomial~\cite{Akbari2013},
our algorithm to compute the independence polynomial can also be used to compute these graph polynomials.
%Some properties of the independence polynomial, such as the uni-modality, log-concavity, and roots of the polynomial, are well studied in the literature for special classes of graphs~\cite{Levit2005}.
%We hope our algorithms to calculate the independence polynomial can also help further research in these directions.
We show some of the Julia language implementations in Appendix~\ref{sec:technical} and you will find it surprisingly short.
A complete implementation can be found in our Github repository~\cite{GraphTensorNetworks}.
%What we need to do is just defining two operations $\oplus$ and $\odot$ and two special elements $\mymathbb{0}$ and $\mymathbb{1}$.
%The style that we program is called generic programming,
